Towards completely caring 15-minute neighbourhoods

### 2024-10-10 paper/Manuscript.qmd
Submitted to journal with editorial correspondance. With lock file update.

---
title: "Caring 15-minute neighbourhoods"
author:
  - name: Anastasia Soukhov 
    email: soukhoa@mcmaster.ca
    affiliations:
        - name: McMaster University
          department: School of Earth, Environment & Society
          city: Hamilton, Canada
    attributes:
        corresponding: true
  - name: Léa Ravensbergen
    email: ravensbl@mcmaster.ca
    affiliations:
        - name: McMaster University
          department: School of Earth, Environment & Society
          city: Hamilton, Canada
  - name: Lucía Mejía Dorantes
    email: lucia.mejia.dorantes@gmail.com
    affiliations:
        - name: Consultant
          city: Karlsruhe, Germany
  - name: Antonio Páez
    email: paezha@mcmaster.ca
    affiliations:
        - name: McMaster University
          department: School of Earth, Environment & Society
          city: Hamilton, Canada
format: 
  # docx
  elsevier-pdf:
    keep-tex: true
    include-in-header:
      text: |
        \usepackage{graphicx}
        \usepackage{unicode-math}
        \usepackage{hyperref}
        \def\tightlist{}
        \usepackage{setspace}
        \doublespacing
        \usepackage{lineno}
        \linenumbers
    journal:
      #name: "Transportation Research Part A: Policy and Practice"
      formatting: preprint
      model: 3p
      cite-style: authoryear
keywords: ["Accessibility", "Mobility of Care", "Gender", "15-minute neighbourhood", "Care Complete Neighbourhoods"]
bibliography: bibliography.bib
---

@@

The "15-Minute City" concept has been embraced by global leaders to promote human-scale neighbourhoods with transport and land-use designs that support short trips to daily necessities. This paper bridges the 15-Minute City to "Mobility of Care",  a framework that foregrounds travel to care destinations, travel done predominately by women. This focus contrasts the more commonly studied travel to employment and leisure destinations. While the 15-Minute City concept is flexible enough to consider all destination types, gendered examinations are relatively lacking in the literature, and little research to date has focused explicitly on care destinations. This gap is addressed in this paper by identifying which areas in the city of Hamilton, Canada are ‘caring 15-minute neighbourhoods’. To do so, a database of care destinations was compiled to estimate the number (using the cumulative opportunities accessibility measure) and diversity of mix (using the entropy measure) of care destinations within a 15-minute walk from residential parcels. Using data-driven machine learning techniques (Self-Organizing Maps and Decision Trees), neighborhoods were classified into 'caring 15-minute neighbourhood' typologies that are examined across residential socio-economic profiles. Our results suggest that the majority of caring 15-minute neighbourhoods are in the urban core, where more lower-income households reside. In contrast, not caring 15-minute neighbourhoods in higher-income peripheral areas. Areas that make good candidates for urban policy intervention are identified, and the implications of enhancing 15-minute walkable caring access are discussed in relation to gender mainstreaming in transportation planning.

Keywords: Accessibility; 15-Minute Neighborhoods; Chrono-Urbanism; Cities of Proximity; Mobility of Care; Inequality; Gender Gap;

@@

The term 15-Minute City is relatively new but the concept is not, as it is simply a normative cumulative-opportunities accessibility measure [@paez_measuring_2012]. Its newfound fame under the moniker of chronourbanism comes to complement long-standing efforts to foster community and local travel to amenities in neighbourhood planning practice. As recent examples, planned neighbourhoods dominated post-WWII built urban form: the Neighbourhood Unit Concept in the western world [@brodyNeighbourhoodUnitConcept2013] and a parallel concept of the “mikrorayons” neighbourhoods in the Soviet Union  [@kissfazekasCircleParadigms15minute2022]. Along different dimensions, these planning forms have been extensively critiqued [@talenSocialSciencePlanned2017]. In this respect, neighbourhood planning approaches such as the 15-Minute City concept offer an opportunity for bottom-up approaches that leverage co-creation tools and meaningful resident participation to achieve equitable and just neighbourhood forms [@mahmoudCocreationPathwayUrban2021]. How to prescribe equitable urban forms that 'enable' contact with social opportunities (instead of 'engendering it' from the top-down) is still a matter for debate. Questions remain, including: what destinations really matter? What mode and travel time threshold is most appropriate? And who benefits? To be certain, these dimensions are under discussion among proximity-based planners who use accessibility-based tools [@silvaProximitycentredAccessibilityConceptual2023; @silvaRegionalAccessibilityUndermining2022; @guzmanProximityEnoughCritical2024].

From another perspective, feminist geographers, urban planners and other researchers agree that urban structure influences travel, but city planning falls short in planning for a neutral identity. In contrast, cities ought to be planned according to the multiple identities of the inhabitants [@vacchelli_towards_2018; @urbandevelopmentviennaGenderMainstreamingUrban2013]. Researchers have demonstrated how mobility patterns differ according to gender [@lawWomenTransportNew1999; @cresswell_gendered_2008; @levy2013travel; @little_gender_1994; @tronto_toward_1990; @neto_gender_2015]. The causes for this are varied, and related to gendered social norms and cultural factors that play a role in how, where, with whom, what for, and how far we move. Of note is that women tend to travel more frequently and for shorter distance [@roorda_triap_2010; @morency_distance_2011] and have more activities related to care [@ilo_care_2018] no matter at what life stage they are [@garciaromanGenderDifferencesTime2022]. Combined, these findings indicate that the spatial organization of activities has a different impact according to gender, and therefore the 15-Minute City concept should not be gender blind.

This work builds a theoretical bridge between the 15-Minute City and mobility of care concepts to answer questions about what destinations matter and who benefits from proximity planning. The Mobility of Care term was coined in @sanchezdemadariagaMobilityCareIntroducing2013; it refers to all travel required to sustain the needs of a household, such as grocery shopping, escorting children, travelling to health appointments, and running errands.  While decades of research have examined types of household-serving trips (e.g., shopping trips, escorting to school), @sanchezdemadariagaMobilityCareIntroducing2013 was the first to consider them all as one category. In doing so, care-related mobility can be shown to comprise a significant proportion of daily travel, characterizing approximately 30% of adults’ daily trips [@sanchezdemadariagaMobilityCareIntroducing2013; @sanchezdemadariagaMeasuringMobilitiesCare2019; @ravensbergen2023exploratory]. We argue that Mobility of Care is especially relevant to the 15-Minute City, as care trips are necessary for _all_ people and they most often occur at the local level. Moreover, care trips are of particular gendered and socio-economic import, as women, and especially those in lower income households, complete the majority of care trips [@ravensbergen2023exploratory]. Despite care travel's importance in advancing gender equality -and equity more broadly-, mobility of care and access to care destinations have been under-examined relative to employment destinations. This is especially pertinent in accessibility research that has largely focused on employment points of interest e.g., [@farberOntarioLineSocioeconomic2019; @duarteInfluenceJobAccessibility2023; @ryanAccessibilitySpaceTime2023].

@@

In the last decades, the need for more sustainable, healthier, and livable cities has become a prominent concern. Planners and decision-makers have proposed a shift to neighbourhood planning principles centered on proximity to urban functions [@pozoukidou15MinuteCityDecomposing2021]. In this context, the 15-Minute City is now in the public spotlight [@logan_xminute_2022; @moreno_introducing_2021]. The 15-Minute City was introduced in @moreno_introducing_2021 along with four dimensions: density (in terms of people per km$^2$), diversity (including mixed-land use and diversity of people), the temporal and spatial proximity to essential services, and digitalisation [q.v. @cerveroTravelDemand3Ds1997]. However, the framework presented has been criticized within academic and planning circles [e.g., @guzmanProximityEnoughCritical2024; @mouratidisTimeChallenge15minute2024] among other things for shortfalls in terms of addressing pre-existing structural forces and individual characteristics that drive inequalities and influence who benefits or could benefit the most from such an approach [@dimarino15minuteCityConcept2023; @willberg15minuteCityAll2023]. A 15-Minute City for all people is an aspirational goal, but does not fully confront existing mobility and accessibility inequalities in need of redress. Without directed and context-specific solutions, this popular concept risks becoming an empty city branding exercise. [@pozoukidou15MinuteCityDecomposing2021; @gowerPlanningInnovationCity2022].

Reflective of the flexible and aspirational presentation of the 15-Minute City, the concept has been adopted by cities using a diverse range of definitions and tools along with indistinct universal approaches. A trail blazer in this respect was Portland (US) in the Portland Plan [@city_of_portland_20minute_2010] of April 2012. Adopted before the '15-Minute City' concept of today, this plan aimed to foster inclusive city development based on prosperity, education, health, and equity over a 30 year horizon. Central to the plan was the promotion of neighbourhood self-sufficiency and connectivity to city centers and centres of employment. The progress report describes a high-level focus on equitable service delivery to all residents with equity concerning topics related to racial equity and people with disabilities [@portland_government_portland_2017], similarly taking on an "all populations" approach. Subsequently, other cities adopted proximity-based goals using similarly neutral approaches. For example, the 15-Minute City plan announced during the re-election campaign of Paris mayor Anne Hidalgo in 2020 emphasized six key social functions that should be easily accessible from any location [@ville_de_paris_paris_2022]. These locations included: housing, work, health care, groceries, education and leisure. The 15-Minute City concept inspired language in the agendas of other cities in the Western world such as Ottawa, in the Canadian context, who also adopted a 15-Minutes approach in their recent Official Plan [@ville_dottawa_quartier_2021]. @teixeiraClassifying15minuteCities2024's global review finds that the 15-Minute City concept is in early phases of implementation around the world and the diverse range of definitions, strategies, and instruments present a significant knowledge and implementation gap. In other words: the 15-Minute City is aspirational but how do we get there and who will benefit?

@@

In examining how to enable improved quality of life through urban built environment, accessibility measures have become an increasingly popular tool. These measures are a way to quantify the ease of reaching destinations from a given point in space and have been used to examine urban areas through just and sustainable city planning agendas [@vale2023accessibility; @handyAccessibilityIdeaWhose2020]. The 15-Minute City is an amenity-provision neighbourhood planning concept aimed at enabling the creation of urban environments that enhance life quality, making it well-suited for analysis using accessibility methodologies [@guzmanProximityEnoughCritical2024]. Recent works have applied accessibility measures to investigate the 15-Minute City across different geographic scopes. For instance, in Naples, Italy [@gaglioneUrbanAccessibility15minute2022], Barcelona, Spain [@graellsCityCitiesMeasuring2021], Vancouver, Canada [@hosford15minuteCityReach2022], and in urban areas across Europe [@vale2023accessibility]. The "cumulative opportunity" measure has been applied in many 15-Minute City examinations. This measure quantifies how many destinations can be reached from a point in space within a given travel time threshold , pairing well with normative examinations [see @paez_measuring_2012] that use travel time thresholds like x-minute cities [@logan_xminute_2022. Furthermore, the cumulative opportunity measure is widely appreciated for its intuitive computation and popularity among practitioners [@handyAccessibilityIdeaWhose2020; @handyMeasuringAccessibilityExploration1997; @chengInvestigatingWalkingAccessibility2019]. However, accessibility measures other than the cumulative opportunity have also been applied, reflecting the diversity of measures in the literature [@guzmanProximityEnoughCritical2024].

@@

Classification algorithms also show promise as a tool in the assessment of the 15-Minute City, as they have been useful in identifying mobility and spatial typologies within the transportation planning literature. Often, these algorithms take the form of machine learning approaches: in transportation, the use of Self-Organizing Maps (SOM) was pioneered by Delmelle [-@delmelle_spatial_2012] and has been used to group U.S. neighbourhoods by minimizing dissimilarity in attributes [@delmelleDifferentiatingPathwaysNeighborhood2017] and to classify individuals' travel patterns according to the dissimilarity in mobility attributes and uses the Decision Tree algorithm to partition the data into interpretable classifications [@victorianoTimeSpaceMoney2020]. Other approaches include the use of spatially constrained multivariate clustering, to develop urban form typologies related to the 15-Minute City [@burkeGeospatialAnalysisFramework2022] and the use of k-means to analyse travel behaviour and classify metro stations based on mobility patterns[ @ganUnderstandingUrbanMobility2020].

@@

Rather than focusing on _all_ destinations, it is valuable to examine those related to caring activities. This framing offers a feminist perspective on urban functions that matter from a care perspective, and connects well with the 15-Minute City concept. Caring activities, which fulfill the physical, psychological, and emotional needs of others, are among the most essential and fundamental activities in society [@ilo_care_2018]. Yet, they are the most unequal, undervalued, and even devalued activities worldwide. Conventionally, caring activities have been borne on women’s shoulders [@haydenGrandDomesticRevolution1982; @hochschild_second_2012]. According to @ilo_care_2018, women and girls perform more than three-quarters of the total amount of unpaid care worldwide, a gender gap that varies geographically [@RN5]. This unequal share of caring responsibilities leads to multifaceted gendered differences: in career development, profession selection, contract type, pay gap, and time poverty, as recognized by various international organisations [@EIGE2016; @ilo_care_2018]. In terms of spatial and transportation planning, almost one third of daily trips are for care purposes [@sanchezdemadariagaMobilityCareIntroducing2013; @sanchezdemadariagaMeasuringMobilitiesCare2019; @ravensbergen2023exploratory]. From this research motivation, Sánchez de Madariaga coined the term Mobility of Care in 2013 to refer to all travel required to sustain the needs of a household, such as grocery shopping, escorting children, travelling to health appointments, and running errands [@sanchezdemadariagaMobilityCareIntroducing2013]. While an undercurrent of research had examined these unique household-serving trips over the decades, her seminal work was the first to consider all these trips as one category and demonstrate how mobility of care is a significant proportion of daily travel.

The Mobility of Care concept also explicitly integrates inter-sectional equity considerations, a common criticism leveled at the 15-Minute City [@guzmanProximityEnoughCritical2024]. Perhaps unsurprisingly given the gendered division of care work discussed, women have consistently been found to complete more mobility of care trips than men. In one study, mobility of care comprised 32% of women’s daily trips compared to 28% of men’s. While this gap is significant, it was found to be far greater in lower income households where women complete 20% more care trips than men [@ravensbergen2023exploratory]. Sánchez de Madariaga not only shows how important these mobility of care trips are, but also highlights the ways in which “mobility of care is systematically under-represented in any analysis of urban transport” (p. 37). As a product of the masculinst bias in planning, transport surveys and tools often do not directly capture mobility of care, which re-enforces the idea that these trips are not a significant part of daily mobility. In this respect, the feminist perspective of the cities of proximity is still underestimated with only few examples on the topic [@gil_sola_choose_2022; @macintyre_her_2022].

Pairing well with the focus on shorter trips and the potential use of sustainable modes within the 15-Minute City, Mobility of Care trips are more local, shorter-distance and trip-chained. Compared to the trip to work, mobility of care trips are more frequently completed by foot, and less frequently by transit or bicycle [@ravensbergen2023exploratory]. However, little work to date examines walking to care destinations through the Mobility of Care framework. Though there is ample literature that examines walking to care destinations, such as schools (e.g., [@RN7;@RN8;@RN9]) and grocery stores (e.g., [@RN10;@RN11]), they tend to consider singular care destinations in research focusing on walkability. 
In summary, we reviewed the neighborhood planning framework from which the 15-Minute City concept stems and is situated within. We then reviewed tools common and useful in investigating the 15-Minute City. Finally, we introduced the Mobility of Care concept, theoretically linking it to the 15-Minute City to highlight the importance of care destinations for all but especially along gendered and socio-economic lines. The objectives of this work are two-fold, achieved through an empirical examination of Hamilton, Canada, a mid-sized city:

@@

A spatial dataset of care destinations for Hamilton was compiled as detailed in ( _work removed for double-blind review_ ). The dataset includes 14 types of care destinations that were classified into four categories: dependent-centric (e.g., the destinations for child- and elder-centric escorting trips), grocery-centric, health-centric, and errand-centric. Notably, these categories were generated following the travel purpose categories created in the mobility of care research by @sanchezdemadariagaMeasuringMobilitiesCare2019. Care categories, sources of data, and preparation notes are detailed in @tbl-Tbl1. The spatial distribution of destination type are visualised in @fig-Fig3 by their category.

@@

Who currently resides in Hamilton's completely caring 15-Minute neighbhorhoods is also demonstrated to be a somewhat optimistic story. Parcels that provide A+ completely caring access tend to be in DAs that are economically disadvantaged. Economic disadvantage tends to intersect with other identities such as gender [@lightmanMeasuringEconomicExclusion2018]. And as reviewed in this work, all women and especially those from lower income households tend to complete most mobility of care trips [@ravensbergen2023exploratory]. Furthermore, lower-income households tend to also be single-parent households. Broadly, single-parent households are more likely to be time-disadvantaged [@nieuwenhuisSingleparentFamiliesWork2018], and tasked with a higher proportion of care duties [@craigTimeCareComparison2004]. In this way, the most economically disadvantaged groups having A+ complete and caring access is an optimistic finding. However, Hamilton is experiencing gentrification [@ellisyoungWeReJust2018]; rents are rising along the future light rail transit corridor and throughout the city [@vandermerweSpilloverGentrificationMidsized2021; @mayersLightTransitDocumenting2023]. Toronto, Hamilton's larger and higher-rent neighbouring city, is spilling gentrification into Hamilton's downtown core, (re)producing neighbourhoods based on Toronto's middle class identities [@mayersLightTransitDocumenting2023]. In these ways, the lower income residents of Hamilton's A+ neighbourhoods are more likely to be displaced, which is matter of wellbeing and justice. There are now few low-rent choices that provide the same exceptional level of access as the downtown core, hence lower household income residents that currently reside in A+ neighhourhoods will likely displaced in the coming years if current trends continue.

>>>>

---
title: "Towards completely caring 15-minute neighbourhoods"
author:
  - name: Anastasia Soukhov 
    email: soukhoa@mcmaster.ca
    affiliations:
        - name: McMaster University
          department: School of Earth, Environment & Society
          city: Hamilton, Canada
    attributes:
        corresponding: true
  - name: Léa Ravensbergen
    email: ravensbl@mcmaster.ca
    affiliations:
        - name: McMaster University
          department: School of Earth, Environment & Society
          city: Hamilton, Canada
  - name: Lucía Mejía Dorantes
    email: lucia.mejia.dorantes@gmail.com
    affiliations:
        - name: Consultant
          city: Karlsruhe, Germany
  - name: Antonio Páez
    email: paezha@mcmaster.ca
    affiliations:
        - name: McMaster University
          department: School of Earth, Environment & Society
          city: Hamilton, Canada
format: 
  # docx
  elsevier-pdf:
    keep-tex: true
    include-in-header:
      text: |
        \usepackage{graphicx}
        \usepackage{unicode-math}
        \usepackage{hyperref}
        \def\tightlist{}
        \usepackage{setspace}
        \doublespacing
        \usepackage{lineno}
        \linenumbers
    journal:
      #name: "Transportation Research Part A: Policy and Practice"
      formatting: preprint
      model: 3p
      cite-style: authoryear
keywords: ["Accessibility", "Mobility of Care", "Gender", "15-minute neighbourhood", "Care Complete Neighbourhoods"]
bibliography: bibliography.bib
---

@@

The "15-Minute City" concept has been embraced by global leaders to promote human-scale neighbourhoods with transport and land-use designs that support short trips to daily necessities. This paper bridges the 15-Minute City to "Mobility of Care",  a framework that foregrounds travel to care destinations, travel done predominately by women. This focus contrasts the more commonly studied travel to employment and leisure destinations. While the 15-Minute City concept is flexible enough to consider all destination types, gendered examinations are relatively lacking in the literature, and little research to date has focused explicitly on care destinations. This gap is addressed in this paper by identifying which areas in the city of Hamilton, Canada are ‘caring 15-minute neighbourhoods’. To do so, a database of care destinations was compiled to estimate the number (using the cumulative opportunity accessibility measure) and diversity of mix (using the entropy measure) of care destinations within a 15-minute walk from residential parcels. Using data-driven machine learning techniques (Self-Organizing Maps and Decision Trees), neighborhoods were classified into 'caring 15-minute neighbourhood' typologies that are examined across residential socio-economic profiles. Our results suggest that the majority of caring 15-minute neighbourhoods are in the urban core, where more lower-income households reside. In contrast, not caring 15-minute neighbourhoods in higher-income peripheral areas. Areas that make good candidates for urban policy intervention are identified, and the implications of enhancing 15-minute walkable caring access are discussed in relation to gender mainstreaming in transportation planning.

Keywords: 15-Minute City; Chrono-Urbanism; Accessibility; Mobility of Care; Inequality; Gender Gap
<!-- Keywords: 15-Minute City; 15-Minute Neighbourhoods; Chrono-Urbanism; Accessibility; Mobility of Care; Cities of Proximity; Inequality; Gender Gap -->

@@

The term 15-Minute City is relatively new, but the concept is not. Its newfound fame under the moniker of chrono-urbanism comes to complement long-standing efforts to foster community and local travel to amenities in neighbourhood planning practice. As recent examples, planned neighbourhoods dominated post-WWII built urban form: the Neighbourhood Unit Concept in the western world [@brodyNeighbourhoodUnitConcept2013] and a parallel concept of the “mikrorayons” neighbourhoods in the Soviet Union  [@kissfazekasCircleParadigms15minute2022]. Along different dimensions, these planning forms have been extensively critiqued [@talenSocialSciencePlanned2017]. In this respect, neighbourhood planning approaches such as the 15-Minute City concept offer an opportunity for bottom-up approaches that leverage co-creation tools and meaningful resident participation to achieve equitable and just neighbourhood forms [@mahmoudCocreationPathwayUrban2021]. How to prescribe equitable urban forms that 'enable' contact with social opportunities (instead of 'engendering it' from the top-down) is still a matter for debate. Questions remain, including: what destinations really matter? What mode and travel time threshold is most appropriate? And who benefits? To be certain, these dimensions are under discussion among proximity-based planners who use accessibility-based tools [@silvaProximitycentredAccessibilityConceptual2023; @silvaRegionalAccessibilityUndermining2022; @guzmanProximityEnoughCritical2024]. In this way, the 15-Minute City could be understood to be a normative cumulative opportunity accessibility measure [@paez_measuring_2012].

From another perspective, feminist geographers, urban planners and other researchers agree that urban structure influences travel, but city planning falls short in planning for a neutral identity. In contrast, cities ought to be planned according to the multiple identities of the inhabitants [@vacchelli_towards_2018; @urbandevelopmentviennaGenderMainstreamingUrban2013]. Researchers have demonstrated how mobility patterns differ according to gender [@lawWomenTransportNew1999; @cresswell_gendered_2008; @levy2013travel; @little_gender_1994; @tronto_toward_1990; @neto_gender_2015]. The causes for this are varied, and related to gendered social norms and cultural factors that play a role in how, where, with whom, what for, and how far we move. Women tend to travel shorter distances and more frequently [@roorda_triap_2010; @morency_distance_2011], and conduct more activities related to care no matter their stage of life [@ilo_care_2018; @garciaromanGenderDifferencesTime2022]. Combined, these findings indicate that the spatial organization of activities has a different impact according to gender, and therefore the 15-Minute City concept should not be gender blind.

This work builds a theoretical bridge between the 15-Minute City and Mobility of Care concepts to answer questions about what destinations matter and who benefits from proximity planning. The Mobility of Care term was coined in @sanchezdemadariagaMobilityCareIntroducing2013; it refers to all travel required to sustain the needs of a household, such as grocery shopping, escorting children, travelling to health appointments, and running errands.  While decades of research have examined types of household-serving trips (e.g., shopping trips, escorting to school), @sanchezdemadariagaMobilityCareIntroducing2013 was the first to consider them all as one category. In doing so, care-related mobility can be shown to comprise a significant proportion of daily travel, characterizing approximately 30% of adults’ daily trips [@sanchezdemadariagaMobilityCareIntroducing2013; @sanchezdemadariagaMeasuringMobilitiesCare2019; @ravensbergen2023exploratory]. We argue that Mobility of Care is especially relevant to the 15-Minute City, as care trips are necessary for _all_ people and they most often occur at the local level. Moreover, care trips are of particular gendered and socio-economic importance, as women, and especially those within lower income households, complete the majority of care trips [@ravensbergen2023exploratory]. Despite care travel's importance in advancing gender equality -and equity more broadly-, Mobility of Care and access to care destinations have been under-examined relative to employment destinations. This is especially pertinent in accessibility research that has largely focused on employment points of interest e.g., [@farberOntarioLineSocioeconomic2019; @duarteInfluenceJobAccessibility2023; @ryanAccessibilitySpaceTime2023].

@@

In the last decades, the need for more sustainable, healthier, and livable cities has become a prominent concern. Planners and decision-makers have proposed a shift to neighbourhood planning principles centered on proximity to urban functions [@pozoukidou15MinuteCityDecomposing2021]. In this context, the 15-Minute City is now in the public spotlight [@logan_xminute_2022; @moreno_introducing_2021]. The 15-Minute City was introduced in @moreno_introducing_2021 along with four dimensions: density (in terms of people per km$^2$), diversity (including mixed-land use and diversity of people), the temporal and spatial proximity to essential services, and digitalisation [related concepts in @cerveroTravelDemand3Ds1997]. However, the framework presented has been criticized within academic and planning circles [e.g., @guzmanProximityEnoughCritical2024; @mouratidisTimeChallenge15minute2024] among other things for shortfalls in terms of addressing pre-existing structural forces and individual characteristics that drive inequalities and influence who benefits or could benefit the most from such an approach [@dimarino15minuteCityConcept2023; @willberg15minuteCityAll2023]. A 15-Minute City for all people is an aspirational goal, but does not fully confront existing mobility and accessibility inequalities in need of redress. Without directed and context-specific solutions, this popular concept risks becoming an empty city branding exercise. [@pozoukidou15MinuteCityDecomposing2021; @gowerPlanningInnovationCity2022].

Reflective of the flexible and aspirational presentation of the 15-Minute City, the concept has been adopted by cities using a diverse range of definitions and tools along with indistinct universal approaches. A trail blazer in this respect was Portland (U.S.) in the Portland Plan [@city_of_portland_20minute_2010] of April 2012. Adopted before the '15-Minute City' concept of today, this plan aimed to foster inclusive city development based on prosperity, education, health, and equity over a 30 year horizon. Central to the plan was the promotion of neighbourhood self-sufficiency and connectivity to city centers and centres of employment. The progress report describes a high-level focus on equitable service delivery to all residents with equity concerning topics related to racial equity and people with disabilities [@portland_government_portland_2017], similarly taking on an "all populations" approach. Subsequently, other cities adopted proximity-based goals using similarly neutral approaches. For example, the 15-Minute City plan announced during the re-election campaign of Paris mayor Anne Hidalgo in 2020 emphasized six key social functions that should be easily accessible from any location [@ville_de_paris_paris_2022]. These locations included: housing, work, health care, groceries, education and leisure. The 15-Minute City concept inspired language in the agendas of other cities in the Western world such as Ottawa, in the Canadian context, who also adopted a 15-Minutes approach in their recent Official Plan [@ville_dottawa_quartier_2021]. @teixeiraClassifying15minuteCities2024's global review finds that the 15-Minute City concept is in early phases of implementation around the world and the diverse range of definitions, strategies, and instruments present a significant knowledge and implementation gap. In other words: the 15-Minute City is aspirational but how do we get there and who will benefit?

@@

In examining how to enable improved quality of life through urban built environment, accessibility measures have become an increasingly popular tool. These measures are a way to quantify the ease of reaching destinations from a given point in space and have been used to examine urban areas through just and sustainable city planning agendas [@vale2023accessibility; @handyAccessibilityIdeaWhose2020]. The 15-Minute City is an amenity-provision neighbourhood planning concept aimed at enabling the creation of urban environments that enhance life quality, making it well-suited for analysis using accessibility methodologies [@guzmanProximityEnoughCritical2024]. Recent works have applied accessibility measures to investigate the 15-Minute City across different geographic scopes. For instance, in Naples, Italy [@gaglioneUrbanAccessibility15minute2022], Barcelona, Spain [@graellsCityCitiesMeasuring2021], Vancouver, Canada [@hosford15minuteCityReach2022], and in urban areas across Europe [@vale2023accessibility]. The "cumulative opportunity" measure has been applied in many 15-Minute City examinations. This measure quantifies how many destinations can be reached from a point in space within a given travel time threshold, pairing well with normative examinations [see @paez_measuring_2012] that use travel time thresholds like x-minute cities [@logan_xminute_2022]. Furthermore, the cumulative opportunity measure is widely appreciated for its intuitive computation and popularity among practitioners [@handyAccessibilityIdeaWhose2020; @handyMeasuringAccessibilityExploration1997; @chengInvestigatingWalkingAccessibility2019]. However, accessibility measures other than the cumulative opportunity have also been applied, reflecting the diversity of measures in the literature [@guzmanProximityEnoughCritical2024].

@@

Classification algorithms also show promise as a tool in the assessment of the 15-Minute City, as they have been useful in identifying mobility and spatial typologies within the transportation planning literature. Often, these algorithms take the form of machine learning approaches: in transportation, the use of Self-Organizing Maps (SOM) was pioneered by Delmelle [-@delmelle_spatial_2012]. It has been used to group U.S. neighbourhoods by minimizing dissimilarity in attributes [@delmelleDifferentiatingPathwaysNeighborhood2017]. SOM has also been used to classify individuals' travel patterns according to the dissimilarity in mobility attributes along with the Decision Tree algorithm to partition the data into interpretable classifications [@victorianoTimeSpaceMoney2020]. Other approaches include the use of spatially constrained multivariate clustering, to develop urban form typologies related to the 15-Minute City [@burkeGeospatialAnalysisFramework2022] and the use of k-means to analyse travel behaviour and classify metro stations based on mobility patterns [@ganUnderstandingUrbanMobility2020].

@@

Rather than focusing on _all_ destinations, it is valuable to examine those related to caring activities. This framing offers a feminist perspective on urban functions that matter from a care perspective, and connects well with the 15-Minute City concept. Caring activities, which fulfill the physical, psychological, and emotional needs of others, are among the most essential and fundamental activities in society [@ilo_care_2018]. Yet, they are the most unequal, undervalued, and even devalued activities worldwide. Conventionally, caring activities have been borne on women’s shoulders [@haydenGrandDomesticRevolution1982; @hochschild_second_2012]. According to @ilo_care_2018, women and girls perform more than three-quarters of the total amount of unpaid care worldwide, a gender gap that varies geographically [@RN5]. This unequal share of caring responsibilities leads to multifaceted gendered differences: in career development, profession selection, contract type, pay gap, and time poverty, as recognized by various international organisations [@EIGE2016; @ilo_care_2018]. In terms of spatial and transportation planning, almost one third of daily trips are for care purposes [@sanchezdemadariagaMobilityCareIntroducing2013; @sanchezdemadariagaMeasuringMobilitiesCare2019; @ravensbergen2023exploratory]. From this research motivation, Sánchez de Madariaga coined the term Mobility of Care in 2013 to refer to all travel required to sustain the needs of a household, such as grocery shopping, escorting children, travelling to health appointments, and running errands [@sanchezdemadariagaMobilityCareIntroducing2013]. While an undercurrent of research had examined these unique household-serving trips over the decades, her seminal work was the first to consider all these trips as one category and demonstrate how Mobility of Care is a significant proportion of daily travel.

The Mobility of Care concept also explicitly integrates inter-sectional equity considerations, a common criticism leveled at the 15-Minute City [@guzmanProximityEnoughCritical2024]. Perhaps unsurprisingly given the gendered division of care work discussed, women have consistently been found to complete more Mobility of Care trips than men. In one study, Mobility of Care comprised 32% of women’s daily trips compared to 28% of men’s. While this gap is significant, it was found to be far greater in lower income households where women complete 20% more care trips than men [@ravensbergen2023exploratory]. Sánchez de Madariaga not only shows how important these Mobility of Care trips are, but also highlights the ways in which “Mobility of Care is systematically under-represented in any analysis of urban transport” (p. 37). As a product of the masculinst bias in planning, transport surveys and tools often do not directly capture Mobility of Care, which re-enforces the idea that these trips are not a significant part of daily mobility. In this respect, the feminist perspective of the cities of proximity is still underestimated with only few examples on the topic [@gil_sola_choose_2022; @macintyre_her_2022].

Pairing well with the focus on shorter trips and the potential use of sustainable modes within the 15-Minute City, Mobility of Care trips are more local, shorter-distance and trip-chained. Compared to the trip to work, Mobility of Care trips are more frequently completed by foot, and less frequently by transit or bicycle [@ravensbergen2023exploratory]. However, little work to date examines walking to care destinations through the Mobility of Care framework. Though there is ample literature that examines walking to care destinations, such as schools (e.g., [@RN7;@RN8;@RN9]) and grocery stores (e.g., [@RN10;@RN11]), they tend to consider singular care destinations in research focusing on walkability. 
In summary, we reviewed the neighborhood planning framework from which the 15-Minute City concept stems and is situated within. We then reviewed tools common and useful in investigating the 15-Minute City. Finally, we introduced the Mobility of Care concept, theoretically linking it to the 15-Minute City to highlight the importance of care destinations for all but especially along gendered and socio-economic lines. The objectives of this work are two-fold, achieved through an empirical examination of Hamilton, Canada, a mid-sized city:

@@

A spatial dataset of care destinations for Hamilton was compiled as detailed in the forthcoming work of @soukhovforthcoming2024. The dataset includes 14 types of care destinations that were classified into four categories: dependent-centric (e.g., the destinations for child- and elder-centric escorting trips), grocery-centric, health-centric, and errand-centric. Notably, these categories were generated following the travel purpose categories created in the Mobility of Care research by @sanchezdemadariagaMeasuringMobilitiesCare2019. Care categories, sources of data, and descriptive notes are detailed in @tbl-Tbl1. The spatial distribution of destination type are visualised in @fig-Fig3 by their category.

@@

Who currently resides in Hamilton's completely caring 15-Minute neighbhorhoods is also demonstrated to be a somewhat optimistic story. Parcels that provide A+ completely caring access tend to be in DAs that are economically disadvantaged. Economic disadvantage tends to intersect with other identities such as gender [@lightmanMeasuringEconomicExclusion2018]. And as reviewed in this work, all women and especially those from lower income households tend to complete most Mobility of Care trips [@ravensbergen2023exploratory]. Furthermore, lower-income households tend to also be single-parent households. Broadly, single-parent households are more likely to be time-disadvantaged [@nieuwenhuisSingleparentFamiliesWork2018], and tasked with a higher proportion of care duties [@craigTimeCareComparison2004]. In this way, the most economically disadvantaged groups having A+ complete and caring access is an optimistic finding. However, Hamilton is experiencing gentrification [@ellisyoungWeReJust2018]; rents are rising along the future light rail transit corridor and throughout the city [@vandermerweSpilloverGentrificationMidsized2021; @mayersLightTransitDocumenting2023]. Toronto, Hamilton's larger and higher-rent neighbouring city, is spilling gentrification into Hamilton's downtown core, (re)producing neighbourhoods based on Toronto's middle class identities [@mayersLightTransitDocumenting2023]. In these ways, the lower income residents of Hamilton's A+ neighbourhoods are more likely to be displaced, which is matter of wellbeing and justice. There are now few low-rent choices that provide the same exceptional level of access as the downtown core, hence lower household income residents that currently reside in A+ neighhourhoods will likely displaced in the coming years if current trends continue.
